Discussion Draft

### 2020-04-05 doc/Discussion.rmd
more additions / clean up of reversed hetC section

The direction of heterochiasmy under the **two locus modfier model** is dependant on female specific recombination modifer supresses or increases recombination, if the modifer is linked to the driving locus, and the meiotic stage (reduction or equalatoral division) that the driver acts (see Table 1 of @brandvain2012scrambling).

@@

Under the **spindle based** model, faster male evolution of the genome wide recombination rates is primarily driven the strength of the spindle assembly checkpoint (SAC) acting on the chromosome structure at metaphase alignment. We refer the reader to [@laneKauppi2019] for more details on the sAC pathway.

The transition from metaphase to anaphase is irreversible,

so all bivalents in the cell have to be aligned correctly to prevent anueploidy.

Unstable bivalents are detected on the MI spindle, the SAC delays the cell cycle which potentially allows time for lose bivalents to correctly attach to the MT spindle. If the 'unattached' persists apoptosis will be triggered. Genetic variants which increase the number of unstable bivalents, will be selected against in favor for genetic variants causing faster stabilization of bivalents on the MI spindle.

Sperm have lower anueploidy rates compared to oocytes (other lines of evidence) indicating that the 'strength' of the SAC varies between oogenesis and spermatogenesis. In sperm single unpaired bivalents can trigger the SAC. While in females more oocytes progress through meiosis even with unpaired bivalents (refereed to as leaky SAC).

These differences in SAC strength have been connected to conserved features of gametogenesis; centrosome spindle and cell volume (cite) which result in differing morphology of the metaphase microtubule spindle.

@@

Directional selection could be the result in spermatogenesis if the fitness landscape for 'chromosome structure' is altered via genetic architecture changes or GxE effects. (in oogenesis the leaky SAC has weak selection coefficient)

>>>>

The direction of heterochiasmy under the **two locus modfier model** is dependent on female specific recombination modifier suppresses or increases recombination, if the modifier is linked to the driving locus, and the meiotic stage (reduction or equatorial division) that the driver acts (see Table 1 of @brandvain2012scrambling).

@@

Under the **spindle based** model, faster male evolution of the genome wide recombination rates is primarily driven the strength of the spindle assembly checkpoint (SAC) acting on the chromosome structure at metaphase alignment. We refer the reader to [@laneKauppi2019] for more details on the SAC pathway.

The transition from metaphase to anaphase is irreversible, so prevention of annueploidy requires that all bivalents be correctly aligned before meiosis can progress (cite). Unstable bivalents detected on the MI spindle, cause the SAC to delay anaphase and allow time for the bivalent to stabilize / connect to the microtubule spindle. If the 'unattached' / unstable bivalents fail to attach, the SAC triggers apoptosis. Genetic variants which increase the number of unstable bivalents, will be selected against in favor for genetic variants causing faster stabilization of bivalents on the MI spindle.

Multiple lines of evidence indicate that the effectiveness of SAC in spermatogenesis is greater compared to oogenesis (LaneKauppi, cite). The different the detection of achiasmated bivalents (), the rates of annuploidy in mature gametes, and stronger infertility phenotypes in mutants of key SAC genes (cite).

These differences in SAC strength have been connected to conserved features of gametogenesis; centrosome spindle and cell volume (cite).

-- Supporting this stage/cell physiology based mechanism  influencing broad sexual dimorphic phenotypes in recombination landscape. (leading to distinct evolutionary trajectories) **directional selection** in males and relaxed selection in females.

@@

<Directional selection could be the result in spermatogenesis if the fitness landscape for 'chromosome structure' is altered via genetic architecture changes or GxE effects. (in oogenesis the leaky SAC has weak selection coefficient)>
